---
title: "22. Reliability analysis"
---

::: {.callout-note appearance="minimal" icon=false}
Language: **R** (Quarto) - [Python version](python.ipynb)
:::

Ported from
[`reliability_analysis.py`](https://github.com/USACE-RMC/Numerics-Python-Examples/blob/main/examples/reliability_analysis.py)
in the USACE-RMC Numerics-Python-Examples repository (0BSD licensed). The upstream
script drives the C# `Numerics.dll` through pythonnet; this version uses
`bestfitr`, whose compiled core is a validated C++ port of the same library. The
Python version of this example uses the same core and prints the same numbers.

This example runs a Monte Carlo reliability analysis of the classic
resistance-minus-load limit state. A component fails when the load S it carries
exceeds the resistance R it can supply, so the limit-state function is g = R - S
and failure is the event g <= 0. Simulating both random variables many times
estimates the failure probability Pf directly, and the reliability index
beta = -&Phi;<sup>-1</sup>(Pf) restates that probability on a standard-normal scale.

## What you'll learn

- How to model a log-normal resistance and a Normal load with `bestfitr`.
- How to estimate a failure probability and reliability index by Monte Carlo simulation.
- How the port's `LnNormal` family replaces the upstream base-e `LogNormal` workaround.

## Setup

The upstream script spends about forty lines resolving `Numerics.dll` from the
NuGet cache and loading the CoreCLR runtime before it can start. None of that
exists here.

```{r}
library(bestfitr)
```

## Define the random variables

Resistance R is log-normal with a physical-space mean of 500 and standard
deviation of 120. Load S is Normal with mean 380 and standard deviation 90.

The upstream script has to work for its log-normal: the C# `LogNormal` class is
base-10 and takes log-space parameters, so the script converts the physical mean
and standard deviation to (mu_ln, sigma_ln) by hand and then mutates
`resistance_dist.Base = e`. That `.Base` mutation is not exposed by this port.
Instead the port ships `LnNormal`, the library's base-e log-normal family, and
`LnNormal` is parameterized in real space: you pass the physical mean and
standard deviation directly, and the class performs the same moment conversion
internally (its `DirectMethodOfMoments` is exactly the algebra the upstream
script writes out). The whole construction collapses to one line.

To show the two parameterizations agree, the chunk below repeats the upstream
hand conversion and checks it against the fitted family: a log-normal's median
is exp(mu_ln), and the physical moments round-trip.

```{r}
resistance_mean <- 500
resistance_std <- 120
resistance_dist <- distribution("LnNormal", c(resistance_mean, resistance_std))
load_dist <- distribution("Normal", c(380, 90))

# The upstream hand conversion, kept only as a cross-check
sigma2_ln <- log1p((resistance_std / resistance_mean)^2)
mu_ln <- log(resistance_mean) - 0.5 * sigma2_ln
cat(sprintf("hand-converted mu_ln = %.6f, sigma_ln = %.6f\n", mu_ln, sqrt(sigma2_ln)))
cat(sprintf("exp(mu_ln)      = %.10f\n", exp(mu_ln)))
cat(sprintf("LnNormal median = %.10f\n", dist_quantile(resistance_dist, 0.5)))

m <- dist_moments(resistance_dist)
cat(sprintf(
  "LnNormal mean = %.6f, sd = %.6f  (round-trips to 500, 120)\n",
  m[["mean"]], m[["sd"]]
))
```

## Monte Carlo simulation

Draw 100,000 samples of each variable with the upstream seeds (42 for
resistance, 43 for load). The port keeps the C# Mersenne Twister bit-exact, so
the streams below are the same numbers the upstream script draws from
`Numerics.dll`, and the Python twin draws them identically.

```{r}
n_samples <- 100000
r <- dist_random(resistance_dist, n_samples, seed = 42)
s <- dist_random(load_dist, n_samples, seed = 43)
g <- r - s
```

## Failure probability and reliability index

Pf is the fraction of samples with g <= 0, and beta = -&Phi;<sup>-1</sup>(Pf).
The upstream script computes beta with `scipy.stats.norm.ppf`; here `qnorm`
supplies the same function, and the Python twin uses the port's own standard
Normal inverse CDF and prints the identical value.

```{r}
n_fail <- sum(g <= 0)
pf <- mean(g <= 0)
beta <- -qnorm(pf)

summary_df <- data.frame(
  Metric = c(
    "Samples", "Failures", "Mean resistance", "Std resistance",
    "Mean load", "Std load", "Pf", "beta"
  ),
  Value = round(c(n_samples, n_fail, mean(r), sd(r), mean(s), sd(s), pf, beta), 6)
)
cat("Reliability analysis summary:\n")
print(summary_df, row.names = FALSE)
```

The first ten simulated samples, as the upstream script prints them:

```{r}
sample_df <- data.frame(
  Resistance = r[1:10], Load = s[1:10], "g = R - S" = g[1:10],
  check.names = FALSE
)
cat("First 10 simulated samples:\n")
print(round(sample_df, 3), row.names = FALSE)
```

## Plots

The upstream script draws one 2x2 figure; here the same four panels are drawn
as separate figures at the site's default size. First, the two input
distributions overlaid:

```{r}
brks <- seq(min(r, s), max(r, s), length.out = 41)
hist(r,
  breaks = brks, freq = FALSE, col = adjustcolor("#6b7f3f", 0.6),
  border = "white", main = "Resistance and load distributions",
  xlab = "Value", ylab = "Density"
)
hist(s,
  breaks = brks, freq = FALSE, col = adjustcolor("#b06a3b", 0.6),
  border = "white", add = TRUE
)
legend("topright",
  legend = c("Resistance", "Load"),
  fill = c("#6b7f3f", "#b06a3b"), bty = "n"
)
```

The limit-state distribution. Everything left of the dashed line is a failure:

```{r}
hist(g,
  breaks = 50, col = "#5b7a8c", border = "white",
  main = "Limit-state distribution", xlab = "g = R - S"
)
abline(v = 0, col = "#b06a3b", lty = 2, lwd = 2)
legend("topright", legend = "Failure boundary g = 0", col = "#b06a3b", lty = 2, bty = "n")
```

The samples in (load, resistance) space. Failures are the points below the
R = S line:

```{r}
fail <- g <= 0
plot(s[!fail], r[!fail],
  pch = 16, cex = 0.2, col = adjustcolor("#6b7f3f", 0.2),
  xlim = range(s, r), ylim = range(s, r),
  xlab = "Load (S)", ylab = "Resistance (R)", main = "Monte Carlo samples"
)
points(s[fail], r[fail], pch = 16, cex = 0.2, col = adjustcolor("#b06a3b", 0.4))
abline(0, 1, lty = 2, lwd = 1.5, col = "#8c8c7a")
legend("topleft",
  legend = c("Safe", "Failure", "R = S"),
  col = c("#6b7f3f", "#b06a3b", "#8c8c7a"),
  pch = c(16, 16, NA), lty = c(NA, NA, 2), bty = "n"
)
```

And the empirical CDF of g. Its value at the dashed line is Pf:

```{r}
plot(sort(g), seq_along(g) / length(g),
  type = "l", lwd = 2, col = "#5b7a8c",
  xlab = "g = R - S", ylab = "Empirical CDF",
  main = "Empirical CDF of the limit state"
)
abline(v = 0, col = "#b06a3b", lty = 2, lwd = 2)
```

## Reproduction check

The upstream file is a standalone script that prints its results at runtime;
the repository does not embed them, so there are no upstream literals to pin
against. The check below therefore asserts two things, stated plainly: internal
consistency (the `LnNormal` physical moments round-trip and the upstream hand
conversion reproduces the internal one), and cross-language identity (the
Python twin asserts exactly the same literals, which only pass because the
seeded draws, Pf, and beta are fully deterministic and bit-identical in both
languages).

| Quantity | Upstream C# | This port | Status |
|----------|------------:|----------:|--------|
| First resistance draw (seed 42) | not embedded upstream | 450.7507979426037 | exact (cross-language) |
| First load draw (seed 43) | not embedded upstream | 271.992998170271 | exact (cross-language) |
| First limit-state value g | not embedded upstream | 178.7577997723327 | exact (cross-language) |
| Failures out of 100,000 | not embedded upstream | 21423 | exact (cross-language) |
| Pf | not embedded upstream | 0.21423 | exact (cross-language) |
| beta | not embedded upstream | 0.7918296694786122 | exact (cross-language) |
| `LnNormal(500, 120)` mean, sd | 500, 120 by construction | `m[["mean"]]`, `m[["sd"]]` | exact (round trip, 1e-12) |

The chunk below fails the render if any value drifts.

```{r}
# Upstream: examples/reliability_analysis.py (standalone script, prints at runtime,
# no embedded outputs). Cross-language identity with the Python twin is the oracle
# here. The values are bit-exact (the fixture suite enforces the seeded stream);
# the comparisons carry a 1e-15 relative tolerance only because R's decimal parser
# can land one ulp away from the written literal.
stopifnot(
  abs(r[1] / 450.7507979426037 - 1) < 1e-15,
  abs(s[1] / 271.992998170271 - 1) < 1e-15,
  abs(g[1] / 178.7577997723327 - 1) < 1e-15,
  n_fail == 21423,
  pf == 0.21423,
  abs(beta / 0.7918296694786122 - 1) < 1e-15
)
# Internal consistency: LnNormal(500, 120) round-trips its physical moments and
# the upstream hand conversion matches the internal one (median = exp(mu_ln)).
stopifnot(
  abs(m[["mean"]] / 500 - 1) < 1e-12,
  abs(m[["sd"]] / 120 - 1) < 1e-12,
  abs(dist_quantile(resistance_dist, 0.5) / exp(mu_ln) - 1) < 1e-14
)
cat("All reproduction checks passed.\n")
```
