---
title: "00. Getting started"
---

::: {.callout-note appearance="minimal" icon=false}
Language: **R** (Quarto) - [Python version](python.ipynb)
:::

Ported from
[`00_getting_started.ipynb`](https://github.com/USACE-RMC/Numerics-Python-Examples/blob/main/notebooks/00_getting_started.ipynb)
in the USACE-RMC Numerics-Python-Examples repository (0BSD licensed). The upstream
notebook drives the C# `Numerics.dll` through pythonnet; this version uses
`corehydror`, whose compiled core is a validated C++ port of the same library, so the
numbers below reproduce the upstream outputs exactly. The Python version of this
example uses the same core and prints the same numbers.

## What you'll learn

- How to install `corehydror` and check the installation.
- How to create a distribution and compute basic statistics.
- What of the upstream setup you can skip entirely (spoiler: all of it).

## Install

The upstream notebook spends its first four steps on .NET plumbing: installing a
.NET runtime, obtaining `Numerics.dll` from NuGet, loading the CoreCLR runtime, and
resolving the DLL path. None of that exists here. `corehydror` ships the ported
library as a regular compiled R package:

```r
install.packages("corehydror")
```

There is no runtime to select, no DLL to resolve, and results are identical on
macOS, Linux, and Windows.

```{r}
library(corehydror)
```

## Basic example: Normal distribution

Create a Normal distribution and compute basic statistics. The upstream code is
`Normal(100, 15)` with property access (`dist.Mean`, `dist.StandardDeviation`, ...);
here every family is constructed by name through `distribution()`, and the moments
come back as one named vector.

```{r}
dist <- distribution("Normal", c(100, 15))
m <- dist_moments(dist)

cat("Mean:", m[["mean"]], "\n")
cat("Standard Deviation:", m[["sd"]], "\n")
cat("Variance:", m[["sd"]]^2, "\n")
cat("Skewness:", m[["skewness"]], "\n")
# Numerics reports Pearson kurtosis (raw fourth-moment form): 3.0 for a Normal.
# The "excess kurtosis" convention is Pearson - 3.
cat("Kurtosis:", m[["kurtosis"]], "\n")

cat("\nSmall sanity check\n")
cat(dist_cdf(distribution("Normal", c(0, 1)), 0), "~= 0.5\n")
```

## No interop plumbing

The upstream notebook needs `System.Array[Double]` conversions for data and
delegate wrappers to pass functions into the C# samplers. With `corehydror`, plain
numeric vectors work everywhere:

```{r}
data <- c(12.0, 15.5, 14.2, 16.8, 13.9)

# Log-likelihood of the data under Normal(mean = 14, sd = 2), evaluated in the core
ll <- dist_log_likelihood(distribution("Normal", c(14, 2)), data)
cat(sprintf("log-likelihood: %.6f\n", ll))
```

## A first taste of seeded reproducibility

The port keeps the C# Mersenne Twister bit-exact, so seeded draws match the
upstream library (and the Python package) exactly. Later examples lean on this
heavily.

```{r}
dist_random(dist, 5, seed = 123)
```

## Reproduction check

Values printed by the upstream notebook (run against the real C# library), compared
with this port. Everything here is deterministic, so the match is exact.

| Quantity | Upstream C# | This port | Status |
|----------|------------:|----------:|--------|
| `Normal(100,15).Mean` | 100.0 | `m[["mean"]]` | exact |
| `Normal(100,15).StandardDeviation` | 15.0 | `m[["sd"]]` | exact |
| `Normal(100,15).Variance` | 225.0 | `m[["sd"]]^2` | exact |
| `Normal(100,15).Skewness` | 0.0 | `m[["skewness"]]` | exact |
| `Normal(100,15).Kurtosis` | 3.0 | `m[["kurtosis"]]` | exact |
| `Normal(0,1).CDF(0)` | 0.5 | `dist_cdf(., 0)` | exact |

The chunk below fails the render if any value drifts. The final check asserts the
cross-language guarantee: the seeded draws above are the exact values the Python
notebook prints.

```{r}
# Upstream reference values: notebook 00_getting_started.ipynb, cell 9 output.
stopifnot(
  m[["mean"]] == 100.0,
  m[["sd"]] == 15.0,
  m[["sd"]]^2 == 225.0,
  m[["skewness"]] == 0.0,
  m[["kurtosis"]] == 3.0,
  dist_cdf(distribution("Normal", c(0, 1)), 0) == 0.5
)

# First seeded draw, identical across R, Python, and the C# library. The draw itself
# is bit-exact (the fixture suite enforces that); the comparison here carries a
# 1e-15 relative tolerance only because R's decimal parser can land one ulp away
# from the written literal.
draw1 <- dist_random(dist, 5, seed = 123)[1]
stopifnot(abs(draw1 / 107.71408449723363 - 1) < 1e-15)
cat("All reproduction checks passed.\n")
```
